---
title: "Population Scale Analysis"
author: "Jessica PID: A15647602"
format: gfm
editor: visual
editor_options: 
  chunk_output_type: inline
---

### Section 4: Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

Answer: The sample size for each genotype is A/A = 108 samples, A/G = 233 samples, and G/G = 121 samples. The median expression levels for each genotype is A/A = 31.25, A/G = 25.06, and G/G = 20.07.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
library(plyr)
table(expr$geno)
exp_meds <- ddply(expr, .(geno), summarise, med = median(exp))
exp_meds

```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Answer. From observing the expression levels of A/A and G/G you can infer that the G/G genotype SNP lowers the expression of the ORMDL3 gene. Additionally, the A/A genotype induces a higher expression of the ORMDL3 gene.

```{r}
library(ggplot2)

ggplot(expr, aes(x = geno, y = exp, fill = geno)) + 
  geom_boxplot(notch = TRUE) + 
  geom_text(data = exp_meds, 
            aes(x = geno, y = med, label = med), size = 3, vjust = -1.5)
```
