---
title: "Class 15: Pertussis miniproject"
author: "Jessica PID: A15647602"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: inline
---

Pertussis, aka Whooping Cough, is a highly contagious lung infection caused by *B. Pertussis*

The CDC tracks Pertussis case numbers and they can be accessed [here](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html)

We need to "scrape" this data so we do stuff with it in R. Let's try the **datapaste** package to do this. 

```{r}
cdc <- data.frame(
          Year = c(
           1922L,1923L,1924L,1925L,
           1926L,1927L,1928L,1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,1936L,
           1937L,1938L,1939L,1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,1947L,
           1948L,1949L,1950L,1951L,1952L,
           1953L,1954L,1955L,1956L,1957L,1958L,
           1959L,1960L,1961L,1962L,1963L,
           1964L,1965L,1966L,1967L,1968L,1969L,
           1970L,1971L,1972L,1973L,1974L,
           1975L,1976L,1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,1984L,1985L,
           1986L,1987L,1988L,1989L,1990L,
           1991L,1992L,1993L,1994L,1995L,1996L,
           1997L,1998L,1999L,2000L,2001L,
           2002L,2003L,2004L,2005L,2006L,2007L,
           2008L,2009L,2010L,2011L,2012L,
           2013L,2014L,2015L,2016L,2017L,2018L,
           2019L,2020L,2021L,2022L,2024L),
         cases = c(
           107473,164191,165418,152003,
           202210,181411,161799,197371,
           166914,172559,215343,179135,265269,
           180518,147237,214652,227319,103188,
           183866,222202,191383,191890,109873,
           133792,109860,156517,74715,69479,
           120718,68687,45030,37129,60886,
           62786,31732,28295,32148,40005,
           14809,11468,17749,17135,13005,6799,
           7717,9718,4810,3285,4249,3036,
           3287,1759,2402,1738,1010,2177,2063,
           1623,1730,1248,1895,2463,2276,
           3589,4195,2823,3450,4157,4570,
           2719,4083,6586,4617,5137,7796,6564,
           7405,7298,7867,7580,9771,11647,
           25827,25616,15632,10454,13278,
           16858,27550,18719,48277,28639,32971,
           20762,17972,18975,15609,18617,
           6124,2116,3044, 23544)
       )
```

```{r}
library(ggplot2)

baseplot <- ggplot(cdc) + 
  aes(y = cases, x = Year) + 
  geom_point() + 
  geom_line()

baseplot

```

Let's add the date of the wP vaccine roll out completion (1946) and the switch to the new aP vaccine (1996).

```{r}
baseplot + 
  theme_bw() +
  geom_vline(xintercept = 1946, col = "blue") +
  geom_vline(xintercept = 1996, col = "red") +
  geom_vline(xintercept = 2020, col = "gray") +
  geom_vline(xintercept = 2004, col = "black")
  
```

## CMI-PB (Computational Models of Immunity - Pertussis Boost)

This project collects and makes freely available data about the immune response to Pertussis vaccination.

You can access the data via an API which returns JSON format (key:value pairs)

We can ise the **jsonlite** package and its`read_json()` function.

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5/subject", simplifyVector = TRUE)
```


```{r}
head(subject)
```

Q. How many subjects do we have?

```{r}
nrow(subject)
```

Q. How many male/femal do we have?

```{r}
table(subject$biological_sex)
```

Q. How many wP and aP do we have?

```{r}
table(subject$infancy_vac)

```

Q. Breakdown of biological sex and race?

```{r}
table(subject$race, subject$biological_sex)
```

Q. Does this breakdown reflect the US population?
A. NO

```{r}
table(subject$dataset)
```

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5/specimen", simplifyVector = TRUE)
head(specimen)
```

```{r}
ab_titer <- read_json("http://cmi-pb.org/api/v5/plasma_ab_titer", simplifyVector = TRUE)
head(ab_titer)
```

We want to merge or "join" these tables so we can have all of the info we need about a given antibody measurement.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
```

```{r}
head(meta)
```

Join `ab_titer` with `meta`

```{r}
abdata <- inner_join(ab_titer, meta)
```

```{r}
head(abdata)
```

```{r}
nrow(abdata)
```

```{r}
table(abdata$isotype)
```

```{r}
table(abdata$antigen)
```

Let's begin with IgG

```{r}
igg <- filter(abdata, isotype == "IgG")
head(igg)
```

Make a boxplot of IgG antigen levels - this will be a plot of MFI vs. antigen

```{r}
ggplot(igg) + 
  aes(x = MFI_normalised, y = antigen, 
      col= infancy_vac) + 
  geom_boxplot() 
```

```{r}
ggplot(igg) + 
  aes(x = MFI_normalised, y = antigen) + 
  geom_boxplot() + 
  facet_wrap(~infancy_vac)
```

Ideally I would like to see how these Ab levels change over time relative to the booster shot. 

```{r}
table(abdata$visit)
```

```{r}
ggplot(igg) + 
  aes(x = MFI_normalised, y = antigen, 
      col = infancy_vac) + 
  geom_boxplot() + 
  facet_wrap(~visit)
```

```{r}
table(abdata$dataset)
```


```{r}
igg2 <- filter(abdata, isotype == "IgG", antigen == "PT", 
               dataset == "2021_dataset")

ggplot(igg2) + 
  aes(y = MFI_normalised, x = planned_day_relative_to_boost, 
      col= infancy_vac, group = subject_id) + 
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept=0, linetype="dashed") +
  geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```


