---
title: "Class10: Halloween Mini Project"
author: "Jessica PID A15647602"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: inline
---
## 1. Importing Candy Data
```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?
A. There are 85 different candy types.

```{r}
nrow(candy)
```
>Q2. How many fruity candy types are in the dataset?
A. There are 38 fruity candy types. 

```{r}
sum(candy$fruity)
```
## 2. What is your favorite candy?

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
A. Snickers; 76.67378

```{r}
candy["Snickers", ]$winpercent
```

>Q4. What is the winpercent value for “Kit Kat”?
A. 76.7686
>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
A. 49.6535

```{r}
candy["Kit Kat", ]$winpercent
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library(dplyr)
candy |>
  filter(rownames(candy) %in% c("Kit Kat", "Tootsie Roll Snack Bars")) |>
  select(winpercent)
```

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
A. winpercent is on a different scale.

>Q7. What do you think a zero and one represent for the candy$chocolate column?
A. The 1 means that the candy has chocolate in it. 

```{r}
inds <- candy$chocolate== 1
choc.win <- candy[inds,]$winpercent

inds <- candy$fruity== 1
fruit.win <- candy[inds,]$winpercent

summary(choc.win)
summary(fruit.win)

t.test(choc.win, fruit.win)

```

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

>Q9. Is the distribution of winpercent values symmetrical?
A. No it is not symmetrical. 

>Q10. Is the center of the distribution above or below 50%?
A. The center of the distribution is below the 50%. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?
A. The chocolate (60) is ranked higher than the candy (44). 

>Q12. Is this difference statistically significant?
A. Yes, the difference is statistically significant.

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```

## 3. Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?
A. Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

```{r}
inds <- order((candy$winpercent))
head(candy[inds,], n=5)
```

>Q14. What are the top 5 all time favorite candy types out of this set?
A. Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

```{r}
inds2 <- order(desc(candy$winpercent))
head(candy[inds2,], n= 5)
```
>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
#Adding useful colors
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 

```

>Q17. What is the worst ranked chocolate candy?
A. The worst ranked chocolate candy is Sixlets. 
>Q18. What is the best ranked fruity candy?
A. The best ranked fruity candy is Starburst.

## 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=2, max.overlaps = 80)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
A. Tootsie Roll Midgies

```{r}
order <- order((candy$winpercent/candy$pricepercent), decreasing = TRUE)
head(candy[order,c(11,12)], n=1 )
```


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

A. The 5 most expensive candies are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular is Nik L Nip. 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

#Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
library(ggplot2)

ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col() 
```

# Make a lollipop chart of pricepercent

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## 5. Exploring the Correlation structure

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```
>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
A. The fruity and chocolate are anti-correlated.
>Q23. Similarly, what two variables are most positively correlated?
A. Winpercent and bar with chocolate are the most positively correlated.


## 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale. = TRUE)
summary(pca)
pca$rotation[,1]
```

```{r}
#Now we can plot our main PCA score plot of PC1 vs PC2.
plot(pca$x[, 1], pca$x[, 2])
```

```{r}
#We can change the plotting character and add some color:
plot(pca$x[,1:2], col=my_cols, pch=16)
```
 
 We can make a much nicer plot with the ggplot2
 
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```
 
```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
 
```{r}
library(ggrepel)

p + geom_text_repel(size= 2, col=my_cols, max.overlaps = 60)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (pink), other (black)",
       caption="Data from 538")
```
 
 
```{r}
#library(plotly)
#ggplotly(p)
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
A. Fruity candy, hard candy, and pluribus candy(coming from a bag or box of multiple candy). Yes these variable make sense. 
